Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases

### 2025-03-15 endofquarterpaper.qmd
fix some cosmetic issues

In this study, we extend the use of geometric signs as identity markers to investigate cultural groups in the Aurignacian period. The Aurignacian period spans roughly 43 - 30 k BP, and is largely characterized by material culture indicators, such as carinated scrapers, small-flake tools and split-based bone and ivory tools, as well as an ongoing evolutionary shift from Neanderthals to early modern humans [@Chu_Richter_2020]. The Aurignacian can be divided into phases, again mainly distinguished by shifts in material culture; Proto, Early, Evolved, and Late Aurignacian [@Tartar_2015], as can be seen in @fig-aur-table. @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main phases, Aur-P1 (Proto/Early), which corresponds with the time period of 43–37 k BP, and Aur-P2 (Evolved/Late), which corresponds with the time periods 37-32 k BP. @Banks_dErrico_Zilhao_2013 used Bayesian modeling methods to further divide Aur-P1 into the Proto and Early Aurignacian, placing the Proto-Aurignacian at 41.5 - 39.9 k BP, and the early Aurignacian at 39.8 - 37.8 k BP. They found that the Proto-Aurignacian occurs during the Greenland Interstadials (GI) 9 and 10, periods of climatic amelioration, while the Early Aurignacian began with the Henrich Stadial, which is characterized by dry and arid conditions, and ends with GI 8.

@@

kable(aur_table)

@@

pretty_print_dates <- 
  function(x) formatC(x, 
                      decimal.mark=".", 
                      big.mark=",", 
                      digits = 0, 
                      format = "f")
date_min <- signbase_years_cal %>% 
  pull(MedianBP) %>% 
  min()
date_max <- signbase_years_cal %>% 
  pull(MedianBP) %>% 
  max()
date_range <- date_max - date_min
date_min_pretty <- pretty_print_dates(date_min)
date_max_pretty <- pretty_print_dates(date_max)
date_range_pretty <- pretty_print_dates(date_range)
```

@@

trans_unique_data <- trans_unique_data %>% 
  mutate(group = case_when((site_name == "Pod Hradem"|
                              site_name == "Fumane" |
                              site_name == "Geissenklösterle") ~ "1",
                           (site_name == "El Castillo" | 
                              site_name == "Hohle Fels" | 
                              site_name == "Labeko Koba") ~ "2"))

@@

artifact_count_median <- median(artifact_count_df$artifact_count)

@@

object_euc_data <- time_unique_data_full%>%
  dplyr::select(id, nobjects) %>% 
  dist(method = "euclidean")

jac_full <- vegan::vegdist(artifact_data, method = "jaccard", binary = TRUE)

@@

##mantel chronological
chron_df <- signbase_full_clean %>% 
  column_to_rownames("id") %>% 
  dplyr::select(MedianBP) %>% 
  dist( method = "euclidean")

chron_sign_df <- signbase_full_clean %>% 
  column_to_rownames("id") %>% 
  dplyr::select(line:star) %>% 
  vegdist("jaccard")

@@

##chron permanova
perman_chron <- adonis2(jac_full~as.factor(time_unique_data_full$time_period), 
                    method = "jaccard",
                    sqrt.dist = TRUE)
perman_chron <- perman_chron %>% 
    data_frame() %>% 
    slice_head()

@@

Our data comes from SignBase 1.0, an open-access collection of geometric signs on mobile objects from Upper Paleolithic Europe. SignBase 1.0 is a record of objects noting only the presence or absence of each sign, without recording the quantity or sequence of individual signs on a given object. We used the identification and categorization of the sign types provided by SignBase's curators [@dutkiewicz2020]. The data set consists of `r nartifacts` objects found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign types. We excluded signs if they fell under the "other" category, due to the wide range of potential possibilities. We also excluded seven sites from the sample (only one sign present: Willendorf; extreme geographic distance from the other sites: Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave; dated much later than the rest of the assemblages: Grotte De La Princesse Pauline and Šandalja II). These exclusions resulted in a sample of `r nartifacts_clean` objects found at `r nsites_clean` sites in `r ncountries_clean` countries, and `r nsigntypes_clean` sign types used in this study (@fig-site-map).

```{r fig.height= 9, fig.width=10}
#| label: fig-site-map
#| fig-cap: "A: Map of sites containing geometric signs on mobile objects from Upper Paleolithic Europe, sampled from the SignBase dataset, divided by Aurignacian phases. B: "

@@

signbase_full_clean$time_period <- factor(signbase_full_clean$time_period, 
                                            levels = c("transitional", "proto_aurignacian", 
                                                       "early_aurignacian", "evolved_aurignacian"), 
                                            labels = c("Transitional", "Proto-Aurignacian",
                                                       "Early Aurignacian", "Evolved Aurignacian"))

signcount_barplot <- ggplot(signbase_full_clean) +
  aes(x = MedianBP, fill = time_period) +
  geom_bar(width = 150) +
  scale_x_reverse() +
  labs(fill = "Time period") +
  theme(legend.key.size = unit(0.5, "cm")) +
  scale_fill_discrete(guide = "none")

@@

plot_grid(sign_map, signcount_barplot, ncol=1, align = "hv", scale = c(1, 0.8), labels = "AUTO")

@@

The distribution of objects across sites is highly skewed, with the median number of objects as `r artifact_count_median`, but Vogelherd has `r nvogel`. To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites. A Mantel test between a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types indicated that there was no significant correlation between object abundance and sign type distribution (Mantel statistic R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).

Ages of the objects were determined by calibrating the uncalibrated ages provided by SignBase, using the rcarbon package [@Crema_Bevan_2021]. The calibrated radiocarbon dates of the sampled data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years. We assigned each object to a chronological phase following the date ranges recorded in @tbl-aur-table. The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` sign types, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` sign types, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` sign types, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` sign types. A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (Mantel statistic R = `r chron_r`, p = `r chron_p`). This confirms that our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.

@@

To evaluate the strength of the groupings found in the seriation and network analyses, we used modularity and perMANOVA tests. The modularity test is a community structure network detection method, and quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006; @Newman_Girvan_2004]. The results are expressed in a score of 0 to 1, with 0 meaning no significant subgroups, and 1 meaning perfect groupings. PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017]. The results are expressed by an R-statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, and a p-value, which quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance). The modularity test and perMANOVA were performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.

@@

mplot <- function(x, ..., fill_colour = "black", title = NULL, row_labels = NULL, col_labels = NULL, country_colors = NULL) {
  d <- data.frame(
    x = rep(1:ncol(x), each = nrow(x)),
    y = rep(nrow(x):1, ncol(x)),
    z = as.integer(x)
  )

  if (!is.null(country_colors)) {
    d$country_color <- rep(country_colors, ncol(x))
  }

@@

  p <- ggplot(d, aes(x, y)) +
    geom_tile(col = "grey50", linewidth = 0.5, aes(fill = ifelse(z == 1, country_color, "white"))) +
    scale_fill_identity() +  
    coord_equal(expand = FALSE) +
    scale_x_continuous(
      breaks = 1:ncol(x),
      labels = col_labels,
      position = "top"
    ) +
    scale_y_continuous(
      breaks = 1:nrow(x),
      labels = row_labels
    ) +
    theme(
      axis.text = element_text(size = rel(1)),
      axis.ticks = element_blank(),
      legend.position = "none",
      plot.title = element_text(hjust = 1)
    ) +
    labs(x = "", y = "") +
    rotate_x_text(angle = 90, hjust = NULL, vjust = NULL)

@@

increase_focus <- function(x, country = NULL) {
  mcp <- apply(
    x, 
    MARGIN = 1, 
    FUN = \(z) { mean(which(z == 1)) }, 
    simplify = FALSE
  ) |> unlist()
  row_order <- order(mcp, na.last = TRUE)
  x <- x[row_order, ]

  if (!is.null(country)) {
    country <- country[row_order]
  }

@@

  list(matrix = x, country = country)
}

concentrate <- function(x, country = NULL, max_iter = 100) {
  old <- x
  not_identical <- TRUE
  iter <- 0

  while (not_identical && iter < max_iter) {
    result <- increase_focus(old, country)
    new <- result$matrix
    country <- result$country
    not_identical <- !identical(old, new)
    old <- new
    iter <- iter + 1
  }

@@

  list(matrix = new, country = country)
}

produce_clusters <- function(artifact_data, artifact_data_unique, method = "seriation"){
  artifact_data <- artifact_data %>% 
    select_if(~ !is.numeric(.) || sum(.) != 0)
  jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
  dm <- as.matrix(jac)

  if(method == "seriation"){
  artifact_matrix <- as.matrix(artifact_data %>% 
                                 mutate(across(everything(), ~replace(., . > 1, 1))))
  rownames(artifact_matrix) <- rownames(artifact_data)

  countries <- artifact_data_unique$group
  unique_countries <- unique(countries)
  color_palette <-  RColorBrewer::brewer.pal(length(unique_countries), "Set3")
  row_colors <- setNames(color_palette, unique_countries)
  result <- concentrate(artifact_matrix, country = countries)
  concentrated_matrix <- result$matrix
  reordered_countries <- result$country
  countries_colors <- row_colors[countries]
  country_colors <- row_colors[reordered_countries]

  final_plot <- mplot(concentrated_matrix,
    country_colors = country_colors)}

@@

  set.seed(500)
  jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
  dm <- as.matrix(jac)
  modul_matrix <-
    graph_from_adjacency_matrix(1 - dm, mode = "undirected")
  mem <- as_membership(parse_number(group_data$group))
  as_group <- modularity(modul_matrix, membership = mem)

@@

  final_table <- data_frame("modularity" = as_group, 
                            "permanova r" = perman_r,
                            "permanova p" = perman_p,
                            "mantel r" = r_value_distance_correlation, 
                            "mantel p" = p_value_distance_correlation)

@@

trans_network <- function(artifact_data){
    jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
    dm <- as.matrix(jac)
    disim <- 1 - dm
    disim[disim<0.1] <- 0
    disim_net <- network(disim,
                         directed=F,
                         ignore.eval=F,
                         names.eval='weight')
    disim_net %v% 'vertex.names' <- row.names(artifact_data)
    set.seed(500)
    final_plot <- ggraph(disim_net, layout="fr") +
      geom_edge_link(aes(width = weight, 
                         alpha=weight), 
                       edge_colour = "black", 
                       show.legend=F) +
      scale_edge_width(range=c(0.5,1.5)) +
      scale_edge_colour_gradient(low = "#CCCCCC",
                                   high = "#000000") +
      geom_node_point(alpha=1, size=5) +
      geom_node_text(aes(label=row.names(artifact_data)), 
                     size=3, 
                     family= "Arial", 
                     repel=T, max.overlaps = 100) +
      theme(text= element_text(size = 15), 
            panel.background = element_rect(fill = 'white')) +
      border(color = "black", size = 0.5)
    return(final_plot)
    }
```

@@

seriation_trans <- produce_clusters(trans_artifact_data, trans_unique_data, method = "seriation")
seriation_proto <- produce_clusters(proto_artifact_data, proto_unique_data, method = "seriation")
seriation_early <- produce_clusters(early_artifact_data, early_unique_data, method = "seriation")
seriation_evolved <- produce_clusters(evolved_artifact_data, evolved_unique_data, method = "seriation")

plot_grid(seriation_trans,seriation_proto, seriation_early , seriation_evolved, 
          ncol = 2, align = "hv", 
          labels = c("Transitional", "Proto", "Early", "Evolved"), label_size = 15, 
          scale = c(0.85, 1, 1, 1))

@@

For each of the phases, seriation identifies at least two groups, one with a restricted range of sign types, and one with a broader range. The notch and line signs are the most prevalent sign types across all time periods. In the Transitional phase, there is a small range of sign types. The restricted group consists of Fumane, Pod Hradem, and Geissenklösterle, two of which only have a notch, and one of which has a notch and circumnotch. The broader group consists of Labeko Koba, Hohle Fels, and El Castillo. In the Proto-Aurignacian we see an expansion in both the number of sites and range of signs, with no overlap in sites from the Transitional phase, except for Geissenklösterle, which is present in both phases. The restricted group consists of Abri Pataud, Gatzarria, Hohlenstein-Stadel, and Mladeč, three of which only have notches and lines, and one also has a dashline. The broader group consists of Geissenklösterle, Spy, Vogelherd, Grotte du Renne, La Ferrassie, and Les Cottés, where we see a more consistent pattern of a more diverse range of signs, compared to the broader group of the Transitional phase. In the Early Aurignacian, we see appearance of completely new sites with no overlap with the Proto-Aurignacian sites, as well as increased diversity of patterns within the groups. The restricted group consists of Grottes de Fonds-de-Forêt (radnotch and notch), Riparo Bombrini (notch), and Grotte de la Verpillière I (notch and line), The broader group consists of Solutré, Hohle Fels, Castanet, Cellier, Blanchard, and Trou al'Wesse, which contain a wide range of signs in a variety of combinations. Lastly, in the Evolved Aurignacian, we see slightly more consistent groups, again with new sites with no overlap with the Early Aurignacian sites, except for Hohle Fels, which also appears in the Transitional phase. Vogelherd also reappears, having appeared earlier in the Proto-Aurignacian.The restricted group consists of Les Rois, Gargas, La Viña, Sirgenstein Cave, and Vindija Cave, covering the line, notch, and and obline signs. The broader group consists of Trou Magrite, Bockstein-Törle, Vogelherd, and Hohle Fels. Overall, across the four phases, the number of sites in the restricted group tends to be less than or equal to the number of sites in the broader group, with the exception of the final Evolved phase.

@@

```{r fig.height= 10, fig.width=10}
#| label: fig-group-sign
#| fig-cap: "Bar plot of signs by time period, colored by group"

full_data_long <- signbase_sf %>% 
  pivot_longer(cols = line:star) %>% 
  filter(value != 0)

ggplot(full_data_long) +
  aes(x = name, fill = group) +
  geom_bar() +
  rotate_x_text(angle = 90, hjust = NULL, vjust = NULL) + 
  facet_wrap(~time_period, nrow = 4)

@@

```{r}
#| label: tbl-strength-table
#| tbl-cap: "Table of the results per phase for the modularity test and perMANOVA"
#| tbl-cap-location: bottom
##Modularity test
trans_strength <- strength_function(trans_artifact_data, trans_unique_data)
proto_strength <- strength_function(proto_artifact_data, proto_unique_data)
early_strength <- strength_function(early_artifact_data, early_unique_data)
evolved_strength <- strength_function(evolved_artifact_data, evolved_unique_data)

full_table <- rbind(trans_strength, proto_strength, early_strength, evolved_strength) %>% 
  mutate(`time period` = c("Transitional", "Proto-Aurignacian",
                                                       "Early Aurignacian", "Evolved Aurignacian"), .before = 1)

strength_table <- full_table %>% 
  dplyr::select(`time period`, modularity, `permanova r`, `permanova p`)

@@

proto_modularity <- proto_strength %>% 
  dplyr::select(modularity)
evolved_modularity <- evolved_strength %>% 
  dplyr::select(modularity)

trans_perm_r <- trans_strength %>% 
  dplyr::select(`permanova r`)
early_perm_r <- early_strength %>% 
  dplyr::select(`permanova r`)

trans_perm_p <- trans_strength %>% 
  dplyr::select(`permanova p`)
```

@@

```{r}
#| label: tbl-mantel-table
#| tbl-cap: "Table of the results per phase for the Mantel test between a Jaccard matrix of sign type diversity and a geographic distance matrix"
#| tbl-cap-location: bottom
## Mantel test for geography vs signs
mantel_table <- full_table %>% 
  dplyr::select(`time period`, `mantel r`, `mantel p`)

@@

proto_mantel_r <- proto_strength %>% 
  dplyr::select(`mantel r`)

proto_mantel_p <- proto_strength %>% 
  dplyr::select(`mantel p`)

early_mantel_r <- early_strength %>% 
  dplyr::select(`mantel r`)

early_mantel_p <- early_strength %>% 
  dplyr::select(`mantel p`)

>>>>

In this study, we extend the use of geometric signs as identity markers to investigate cultural groups in the Aurignacian period. The Aurignacian period spans roughly 43 - 30 k BP, and is largely characterized by material culture indicators, such as carinated scrapers, small-flake tools and split-based bone and ivory tools, as well as an ongoing evolutionary shift from Neanderthals to early modern humans [@Chu_Richter_2020]. The Aurignacian can be divided into phases which are mainly distinguished by shifts in material culture; Proto, Early, Evolved, and Late Aurignacian [@Tartar_2015], as can be seen in @tbl-aur-table. @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main phases, Aur-P1 (Proto/Early), which corresponds with the time period of 43–37 k BP, and Aur-P2 (Evolved/Late), which corresponds with the time periods 37-32 k BP. @Banks_dErrico_Zilhao_2013 used Bayesian modeling methods to further divide Aur-P1 into the Proto and Early Aurignacian, placing the Proto-Aurignacian at 41.5 - 39.9 k BP, and the early Aurignacian at 39.8 - 37.8 k BP. They found that the Proto-Aurignacian occurs during the Greenland Interstadials (GI) 9 and 10, periods of climatic amelioration, while the Early Aurignacian began with the Henrich Stadial, which is characterized by dry and arid conditions, and ends with GI 8.

@@

kable(aur_table)
```

@@

pretty_print_dates <- 
  function(x) formatC(x, 
                      decimal.mark=".", 
                      big.mark=",", 
                      digits = 0, 
                      format = "f")

date_min <- signbase_years_cal %>% 
  pull(MedianBP) %>% 
  min()
date_max <- signbase_years_cal %>% 
  pull(MedianBP) %>% 
  max()
date_range <- date_max - date_min
date_min_pretty <- pretty_print_dates(date_min)
date_max_pretty <- pretty_print_dates(date_max)
date_range_pretty <- pretty_print_dates(date_range)
```

@@

# BM: a little simpler and less typings using %in% 
trans_unique_data <- trans_unique_data %>% 
  mutate(group = case_when(site_name %in% c("Pod Hradem", "Fumane", "Geissenklösterle") ~ "1",
                           site_name %in% c("El Castillo", "Hohle Fels", "Labeko Koba") ~ "2"))

@@

artifact_count_median <- median(artifact_count_df$artifact_count)
```

@@

# BM: not sure why the id column was also selected here? I moved it to rownames
object_euc_data <- time_unique_data_full %>%
    column_to_rownames("id") %>% 
    dplyr::select(nobjects) %>% 
    dist(method = "euclidean")

# BM: Ensure binary data as input to jaccard distance
artifact_data_bin <- 
artifact_data %>% 
  mutate_all(~ as.numeric(. > 0))

jac_full <- vegan::vegdist(artifact_data_bin, method = "jaccard", binary = TRUE)

@@

## mantel chronological
chron_df <- signbase_full_clean %>% 
  column_to_rownames("id") %>% 
  dplyr::select(MedianBP) %>% 
  dist( method = "euclidean")

chron_sign_df <- signbase_full_clean %>% 
  column_to_rownames("id") %>% 
  dplyr::select(line:star) %>% 
  # BM: Ensure binary data as input to jaccard distance
  mutate_all(~ as.numeric(. > 0)) %>% 
  vegdist("jaccard")

@@

## chron permanova
perman_chron <- adonis2(jac_full ~ as.factor(time_unique_data_full$time_period), 
                    method = "jaccard",
                    sqrt.dist = TRUE)

perman_chron <- perman_chron %>% 
    data_frame() %>% 
    slice_head()

@@

Our data come from SignBase 1.0, an open-access collection of geometric signs on mobile objects from Upper Paleolithic Europe. SignBase 1.0 is a record of objects noting only the presence or absence of each sign on the object, without details of the quantity or sequence of individual signs on a given object. We used the identification of sign types provided by SignBase's curators [@dutkiewicz2020]. The data set consists of `r nartifacts` objects found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign types. We excluded signs if they fell under the "other" category, due to the wide range of potential possibilities. We also excluded seven sites from the sample (only one sign present: Willendorf; extreme geographic distance from the other sites: Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave; dated much later than the rest of the assemblages: Grotte De La Princesse Pauline and Šandalja II). These exclusions resulted in a sample of `r nartifacts_clean` objects found at `r nsites_clean` sites in `r ncountries_clean` countries, and `r nsigntypes_clean` sign types used in this study (@fig-site-map).

```{r fig.height= 9, fig.width=10}
#| label: fig-site-map
#| fig-cap: "A: Map of sites in each Aurignacian phase that contain geometric signs on mobile objects. B: Abundance of signs over time, phases indicated by colours. C. Distribution of object count by site"
#| fig-height: 10

@@

signbase_full_clean$time_period <-
  factor(
    signbase_full_clean$time_period,
    levels = c(
      "transitional",
      "proto_aurignacian",
      "early_aurignacian",
      "evolved_aurignacian"
    ),
    labels = c(
      "Transitional",
      "Proto-Aurignacian",
      "Early Aurignacian",
      "Evolved Aurignacian"
    )
  )

signcount_barplot <- ggplot(signbase_full_clean) +
  aes(x = MedianBP, fill = time_period) +
  geom_bar(width = 150) +
  scale_x_reverse() +
  labs(fill = "Time period") +
  theme(legend.key.size = unit(0.5, "cm")) +
  theme_minimal() +
  scale_fill_discrete(guide = "none") +
  xlab("Years cal BP")

objectcount_barplot <- 
ggplot(artifact_count_df) +
  aes(reorder(site_name, artifact_count),
      artifact_count) +
  geom_col() +
  coord_flip() +
    theme_minimal() +
    xlab("") +
    ylab("Object count")

@@

maps_bar_plots <- 
plot_grid(sign_map,
          plot_grid(signcount_barplot,
                    objectcount_barplot,
                    rel_widths = c(1, 0.5),
                    labels = c("B", "C")), 
          ncol = 1, 
          scale = c(1, 1),
          rel_heights = c(1, 0.5),
          align = "vh", 
          labels = "A",
          axis = "lr")

ggsave(maps_bar_plots, 
      filename =  "map-and_hist-plot.png",
       h = 10, w = 10,
       bg = "white")

knitr::include_graphics("map-and_hist-plot.png")

@@

The distribution of objects across sites is highly skewed, with the median number of objects as `r artifact_count_median`, but Vogelherd has `r nvogel`. To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites. A Mantel test between a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).

Ages of the objects were determined by calibrating the uncalibrated ages provided by SignBase, using the rcarbon package [@Crema_Bevan_2021]. The calibrated radiocarbon dates of the sampled data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years. We assigned each object to a chronological phase following the date ranges recorded in @tbl-aur-table. The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` sign types, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` sign types, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` sign types, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` sign types. A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (R = `r chron_r`, p = `r chron_p`). This confirms that our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.

@@

To evaluate the strength of the groupings found in the seriation and network analyses, we used modularity and perMANOVA tests. The modularity test is a community structure network detection method, and quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006; @Newman_Girvan_2004]. The results are expressed in a score of 0 to 1, with 0 meaning no significant subgroups, and 1 meaning perfect groupings. PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017]. The results are expressed by an R statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, and a p-value, which quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance). The modularity test and perMANOVA were performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.

@@

mplot <- function(x, ..., fill_colour = "black", title = NULL, row_labels = NULL, col_labels = NULL, palette_colors = NULL) {
  d <- data.frame(
    x = rep(1:ncol(x), each = nrow(x)),
    y = rep(nrow(x):1, ncol(x)),
    z = as.integer(x)
  )

  if (!is.null(palette_colors)) {
    d$palette_colors <- rep(palette_colors, ncol(x))
  }

@@

  p <- ggplot(d, aes(x, y)) +
    geom_tile(col = "grey50", 
              linewidth = 0.5, 
              aes(fill = ifelse(z == 1, palette_colors, "white"))) +
    scale_fill_identity() +  
    coord_equal(expand = FALSE) +
    scale_x_continuous(
      breaks = 1:ncol(x),
      labels = col_labels,
      position = "top"
    ) +
    scale_y_continuous(
      breaks = 1:nrow(x),
      labels = row_labels
    ) +
    theme(
      axis.text = element_text(size = rel(1)),
      axis.ticks = element_blank(),
      legend.position = "none",
      plot.title = element_text(hjust = 1)
    ) +
    labs(x = "", y = "") +
    rotate_x_text(angle = 90, hjust = 0, vjust = 0.5)

@@

increase_focus <- function(x, group = NULL) {
  mcp <- apply(
    x, 
    MARGIN = 1, 
    FUN = \(z) { mean(which(z == 1)) }, 
    simplify = FALSE
  ) |> unlist()
  row_order <- order(mcp, na.last = TRUE)
  x <- x[row_order, ]

  if (!is.null(group)) {
    group <- group[row_order]
  }

@@

  list(matrix = x, group = group)
}

concentrate <- function(x, group = NULL, max_iter = 100) {
  old <- x
  not_identical <- TRUE
  iter <- 0

  while (not_identical && iter < max_iter) {
    result <- increase_focus(old, group)
    new <- result$matrix
    group <- result$group
    not_identical <- !identical(old, new)
    old <- new
    iter <- iter + 1
  }

@@

  list(matrix = new, group = group)
}

produce_clusters <- function(artifact_data, artifact_data_unique, method = "seriation", color_palette = c("#8DD3C7", "#FFFFB3")){

  artifact_data <- artifact_data %>% 
    select_if(~ !is.numeric(.) || sum(.) != 0) %>% 
    mutate_all(~ as.numeric(. > 0))

  jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
  dm <- as.matrix(jac)

  if(method == "seriation"){
  artifact_matrix <- as.matrix(artifact_data)
  rownames(artifact_matrix) <- rownames(artifact_data)

  group <- artifact_data_unique$group
  unique_group <- unique(group)
 # color_palette <-  RColorBrewer::brewer.pal(length(unique_group), "Set3")
  row_colors <- setNames(color_palette, unique_group)
  result <- concentrate(artifact_matrix, group = group)
  concentrated_matrix <- result$matrix
  reordered_group <- result$group
  group_colors <- row_colors[reordered_group]

  final_plot <- mplot(concentrated_matrix,
    palette_colors = group_colors)
  }

@@

  set.seed(500)
  jac <- vegdist(artifact_data %>% mutate_all(~ as.numeric(. > 0)), "jaccard", binary = TRUE)
  dm <- as.matrix(jac)
  modul_matrix <-
    graph_from_adjacency_matrix(1 - dm, mode = "undirected")
  mem <- as_membership(parse_number(group_data$group))
  as_group <- modularity(modul_matrix, membership = mem)

@@

  final_table <- data_frame("Modularity" = as_group, 
                            "PerMANOVA r" = perman_r,
                            "PerMANOVA p" = perman_p,
                            "Mantel R" = r_value_distance_correlation, 
                            "Mantel p" = p_value_distance_correlation)

@@

trans_network <- function(artifact_data){
    jac <- vegdist(artifact_data %>%  mutate_all(~ as.numeric(. > 0)), "jaccard", binary = TRUE)
    dm <- as.matrix(jac)
    disim <- 1 - dm
    disim[disim<0.1] <- 0
    disim_net <- network(disim,
                         directed=F,
                         ignore.eval=F,
                         names.eval='weight')
    disim_net %v% 'vertex.names' <- row.names(artifact_data)
    set.seed(500)
    final_plot <- ggraph(disim_net, layout="fr") +
      geom_edge_link(aes(width = weight, 
                         alpha=weight), 
                       edge_colour = "black", 
                       show.legend=F) +
      scale_edge_width(range=c(0.5,1.5)) +
      scale_edge_colour_gradient(low = "#CCCCCC",
                                   high = "#000000") +
      geom_node_point(alpha=1, size=5) +
      geom_node_text(aes(label=row.names(artifact_data)), 
                     size=3, 
                     family= "Arial", 
                     repel=T, max.overlaps = 100) +
      theme(text= element_text(size = 15), 
            panel.background = element_rect(fill = 'white')) +
      border(color = "black", size = 0.5)
    return(final_plot)
    }
```

@@

seriation_trans <- 
  produce_clusters(trans_artifact_data, 
                   trans_unique_data, 
                   method = "seriation")

seriation_proto <- 
  produce_clusters(proto_artifact_data, 
                   proto_unique_data, 
                   # BM: not sure why this one gets the 
                   # colour differently from the others
                   color_palette = c("#FFFFB3", "#8DD3C7"),
                   method = "seriation")

seriation_early <- 
  produce_clusters(early_artifact_data, 
                   early_unique_data, 
                   method = "seriation")

seriation_evolved <- 
  produce_clusters(evolved_artifact_data, 
                   evolved_unique_data, 
                   method = "seriation")

plot_grid(seriation_trans,
          seriation_proto, 
          seriation_early , 
          seriation_evolved, 
          ncol = 2, 
          align = "hv", 
          labels = c("Transitional", "Proto", "Early", "Evolved"), label_size = 15, 
          scale = c(1, 1, 1, 1))

@@

For each of the phases, seriation identifies at least two groups, one with a restricted range of sign types, and one with a broader range. The notch and line signs are the most prevalent sign types across all time periods. In the Transitional phase, there is a small range of sign types. The restricted group consists of Fumane, Pod Hradem, and Geissenklösterle, two of which only have a notch, and one of which has a notch and circumnotch. The broader group consists of Labeko Koba, Hohle Fels, and El Castillo. In the Proto-Aurignacian we see an expansion in both the number of sites and range of signs, with no overlap in sites from the Transitional phase, except for Geissenklösterle, which is present in both phases. The restricted group consists of Abri Pataud, Gatzarria, Hohlenstein-Stadel, and Mladeč, three of which only have notches and lines, and one also has a dashline. The broader group consists of Geissenklösterle, Spy, Vogelherd, Grotte du Renne, La Ferrassie, and Les Cottés, where we see a more consistent pattern of a more diverse range of signs, compared to the broader group of the Transitional phase. In the Early Aurignacian, we see appearance of completely new sites with no overlap with the Proto-Aurignacian sites, as well as increased diversity of patterns within the groups. The restricted group consists of Grottes de Fonds-de-Forêt (radnotch and notch), Riparo Bombrini (notch), and Grotte de la Verpillière I (notch and line), The broader group consists of Solutré, Hohle Fels, Castanet, Cellier, Blanchard, and Trou al'Wesse, which contain a wide range of signs in a variety of combinations. Lastly, in the Evolved Aurignacian, we see slightly more consistent groups, again with new sites with no overlap with the Early Aurignacian sites, except for Hohle Fels, which also appears in the Transitional phase. Vogelherd also reappears, having appeared earlier in the Proto-Aurignacian.The restricted group consists of Les Rois, Gargas, La Viña, Sirgenstein Cave, and Vindija Cave, covering the line, notch, and and obline signs. The broader group consists of Trou Magrite, Bockstein-Törle, Vogelherd, and Hohle Fels. Overall, across the four phases, the number of sites in the restricted group tends to be less than or equal to the number of sites in the broader group, with the exception of the final Evolved phase. <!-- BM: I reckon we need to say something about the signs also, not just the sites -->

@@

@@

```{r}
#| label: tbl-strength-table
#| tbl-cap: "Table of the results per phase for the modularity test and perMANOVA"
#| tbl-cap-location: bottom

## Modularity test
trans_strength <- strength_function(trans_artifact_data, trans_unique_data)
proto_strength <- strength_function(proto_artifact_data, proto_unique_data)
early_strength <- strength_function(early_artifact_data, early_unique_data)
evolved_strength <- strength_function(evolved_artifact_data, evolved_unique_data)

full_table <- rbind(trans_strength, proto_strength, early_strength, evolved_strength) %>% 
  mutate(`Aurignacian phase` = c("Transitional", 
                           "Proto-Aurignacian",
                           "Early Aurignacian", 
                           "Evolved Aurignacian"), 
         .before = 1)

strength_table <- full_table %>% 
  dplyr::select(`Aurignacian phase`, Modularity, `PerMANOVA r`, `PerMANOVA p`) %>% 
  mutate(across(where(is.numeric), ~ round(., 3)))

@@

proto_modularity <- proto_strength %>% 
  dplyr::select(Modularity)
evolved_modularity <- evolved_strength %>% 
  dplyr::select(Modularity)

trans_perm_r <- trans_strength %>% 
  dplyr::select(`PerMANOVA r`)
early_perm_r <- early_strength %>% 
  dplyr::select(`PerMANOVA r`)

trans_perm_p <- trans_strength %>% 
  dplyr::select(`PerMANOVA p`)
```

@@

```{r}
#| label: tbl-mantel-table
#| tbl-cap: "Table of the results per phase for the Mantel test between a Jaccard matrix of sign type diversity and a geographic distance matrix"
#| tbl-cap-location: bottom
## Mantel test for geography vs signs
mantel_table <- full_table %>% 
  dplyr::select(`Aurignacian phase`, `Mantel R`, `Mantel p`)

@@

proto_mantel_r <- proto_strength %>% 
  dplyr::select(`Mantel R`)

proto_mantel_p <- proto_strength %>% 
  dplyr::select(`Mantel p`)

early_mantel_r <- early_strength %>% 
  dplyr::select(`Mantel R`)

early_mantel_p <- early_strength %>% 
  dplyr::select(`Mantel p`)
